## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

```{r}
if (logical statement) {
  do something
} else {
  do something else
}
```

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10

if(a <= 5){
  print(a)
}else{
  print("It's more than 5")
}
```

We can also add else if statements in the middle if we need more options.

```{r}
a <- 8

if(a <= 5){
  print("small")
}else if(a > 5 & a <= 10){
  print("medium")
}else if(a > 10 | a == 7){
  print("large")
}else{
  print("huge")
}
```

Neat! We can also nest conditionals just like we do in Python.

```{r}
a <- 5
if(a / 2 != 0){
  if(a == 5){
    print("a is 5")
  }
}
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for (value in 1:5){
  # paste is a concatenation function
  print(paste(value, "is greater than", value -1))
}
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <- c(1, 1, 1, 1, 1, 1, 1)

# Change the first, third, and sixth elements to the value 6
for (i in c(1, 3, 6)){
  a_vector[i] <- 6
}
a_vector
```

We can loop over more than just lists:

```{r}
mat <- matrix(1, nrow = 2, ncol = 2)

for (row in 1:2){
  for(col in 1:2){
    mat[row, col] <- 0
    print(mat)
  }
}
```

Like Python, we have while loops in R:

```{r}
b <- 0
while(b < 10){
  # increase a by 2
  b <- b + 2
  # print b
  print(b)
}
```

## Functions

When should we use functions?

```{r}
our_function <- function(optional_params){
  # do some stuff
}
```

An example:

```{r}
# Start with a vector of temperatures
test_temp <- c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735)

# Then we want a function that converts from C to F:
c_to_f <- function(dat){
  temp_fahr <- dat*(9/5) + 32
  
  return(temp_fahr)
}

print(c_to_f(26))
```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
data_fahr <- c_to_f(test_temp)
data_fahr
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
test_mat <- matrix(c(1, 1, 1, 2, 2, 2), nrow <- 2)
# 1 is for rows, 2 is for columns
apply(test_mat, 2, sum)
```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_temp_mat <- matrix(c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735), nrow = 2)

apply(test_temp_mat, 1, c_to_f)
```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}
# Given 2 strings, a and b, return a string of the form short+long+short, with the shorter string on the outside and the longer string on the inside. The strings will not be the same length, but they may be empty ( zero length ).
    
short_long_short <- function(a, b) {
    ifelse(nchar(a) < nchar(b), paste0(a, b, a), paste0(b, a, b))
}
    ```
    
```{r}
# After a hard quarter in the office you decide to get some rest on a vacation. So you will book a flight for you and your girlfriend and try to leave all the mess behind you.
# You will need a rental car in order for you to get around in your vacation. The manager of the car rental makes you some good offers.
# Every day you rent the car costs $40. If you rent the car for 7 or more days, you get $50 off your total. Alternatively, if you rent the car for 3 or more days, you get $20 off your total.
# Write a code that gives out the total amount for different days(d).

rental_car_cost <- function(d){
  if (d >= 7){
    d * 40 - 50
  }else if (d >= 3){
    d * 40 - 20
  }else{
    d * 40
  }
  
}
```

```{r}
# You were camping with your friends far away from home, but when it's time to go back, you realize that your fuel is running out and the nearest pump is 50 miles away! You know that on average, your car runs on about 25 miles per gallon. There are 2 gallons left.
# Considering these factors, write a function that tells you if it is possible to get to the pump or not.
# Function should return true if it is possible and false if not.

zero_fuel <- function(distance, mpg, fuel_left) {
  distance <= fuel_left * mpg
}
```

```{r}
# Your task is to make a function that can take any non-negative integer as an argument and return it with its digits in descending order. Essentially, rearrange the digits to create the highest possible number.

descending_order <- function(num){
  as.numeric(paste(sort(unlist(strsplit(as.character(num), NULL)), decreasing = TRUE), collapse = ""))
}
```

```{r}
# Welcome. In this kata, you are asked to square every digit of a number and concatenate them.
# For example, if we run 9119 through the function, 811181 will come out, because 92 is 81 and 12 is 1. (81-1-1-81)
# Example #2: An input of 765 will/should return 493625 because 72 is 49, 62 is 36, and 52 is 25. (49-36-25)

square_digits <- function(num){
  split_num = as.numeric(strsplit(as.character(num), "")[[1]])
  as.integer(paste(split_num^2, collapse = ""))
}
```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}
fizzbuzz <- function(input, mod1, mod2, exp1, exp2) {
  output <- ""
  
  if (input %% mod1 == 0){
    output <- paste(output, exp1)
    }
  if (input %% mod2 == 0){
    output <- paste(output, exp2)
    }
  if (output == ""){
    output <- as.character(input)
    }
  
  return(output)
}

output <- map_chr(c(2, 93, 41, 32, 60, 28, 63, 15, 8), ~ fizzbuzz(.x, 3, 5, "Fizz", "Buzz"))
print(output)
    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    # You may feel more comfortable figuring this out in Python first, then translating into R later.
    
    marioStairs <- function(number){
      stairs = c()
      
      for(i in 1:number){
        for(j in 1:i+1){
          stairs = c(stairs, 'X')
        }
        print(stairs)
        stairs = c()
      }
    }
    ```

```{r}
marioStairs(4)
```

